---
title: "Site selection"
author: "George Roff"
date: "2024-08-13"
html_document:
    theme: null       # Remove theme styling
    self_contained: true
    css: null         # Remove any linked CSS files
    toc: false        # Remove table of contents (if exists in the header)
    includes:         # Prevent any additional content from being included
      before_body: null
      after_body: null
---

style
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<hr>

Comparisons of various approaches to determine optimal placement of restoration areas given an underlying habitat `shp` file using Lizard Island example from [Allen Coral Atlas](https://allencoralatlas.org).

<hr>

### Import habitats 

i) import `shp` files, filter and remove polygons less than target plot hectare size (1ha).

Red inset box = subset map used below (note - spatial data in projected coordinate system 20353)


```{r libraries, message=FALSE, warning=FALSE, cache=TRUE, fig.width=6, fig.height=6}

print("test")

```
